disaggregation: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling

### 2019-12-03 article.Rnw
Minor edit
Formatting for JSS. Refs in title case (used perl script)

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{disaggregation: An \proglang{R} package for Bayesian spatial disaggregation modelling}
\Plaintitle{disaggregation: An R package for Bayesian spatial disaggregation modelling}
\Shorttitle{disaggregation modelling in \proglang{R}}

@@

Disaggregation modelling is unorthodox as the predictions are at a different scale to the response data, i.e. the number of rows in the covariate data is different to that of the response data. The spatial modelling software package, \pkg{INLA}~\cite{rue2009approximate}, has been shown to very useful in a wide variety of circumstances, however it is not flexible enough for the unorthodox nature of the disaggregation problem except in the special case of the linear link function \cite{wilson2017pointless}. Disaggregation models can be implemented in \pkg{TMB}~\cite{tmb2016} with a lot of flexibility, however the data manipulation required to format the model objects and construct the model definition in C++ is non-trivial. The \pkg{disaggregation} package allows this process to be streamlined, to make it easy for the user to run disaggregation models at the expense of some flexibility.

@@

The \pkg{disaggregation} package is built on Template Model Builder (\pkg{TMB})~\cite{tmb2016}, which is a tool for flexibly building complex models based on C++. \pkg{TMB} combines the packages CppAD~\cite{bell2012cppad}, for automatic differentiation in C++, Eigen~\cite{guennebaud2010eigen}, a C++ library for linear algebra, and CHOLMOD~\cite{chen2008algorithm}, a package for efficient computation of sparse matrices. The use of these packages allows an efficient implementation of the automatic Laplace approximation~\cite{skaug2006automatic} with exact derivatives which gives an approximation to the Bayesian posterior. \pkg{TMB} calculates first and second order derivatives of the objective function using automatic differentiation~\cite{griewank2008evaluating}.

@@

%
<<setuppaths, results='hide'>>=
library(raster)

@@

%
<<prepare_data, warning = FALSE>>=
dis_data <- prepare_data(polygon_shapefile = shapes, 
                         covariate_rasters = covariate_stack, 
                         aggregation_raster = population_raster, 
                         mesh.args = list(max.edge = c(0.7, 8), 
                                          cut = 0.05, 
                                          offset = c(1, 2)),
                         id_var ='ID_2',
                         response_var = 'inc',
                         na.action = TRUE,
                         ncores = 8)
@
%

@@

\section{Comparison with MCMC} \label{sec:comparison}

In this section we show performance comparisons between modelling using the Laplace approximation provided by the \pkg{disaggregation} package, based on \pkg{TMB}, and using Markov chain Monte Carlo (MCMC). Given a function to evaluate the probability density of a distribution at any given point in parameter space, MCMC algorithms construct markov chains to generate samples from this distribution. These algorithms are often slow, particularly in high-dimensional settings, as it can take a long time to converge to and effectively sample from the stationary distribution. The density is evaluated at each step, so this problem is compounded when evaluating this density is computationally expensive. In contrast, the \pkg{disaggregation} package approximates uses a Laplace approximation to the posterior to generate posterior samples. This only requires the posterior to be maximised to find the posterior mode and therefore involves relatively few evaluations of the posterior density compared to MCMC techniques, although potentially at the expense of less accurate posterior samples. Here we compare the time and performance of the two techniques.

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS algorithm \cite{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.

@@

The modelling framework is implemented using the Laplace approximation and automatic differentiation within the \pkg{TMB} package. This allows fast, optimised calculations in C++. These disaggregation models are computationally intensive and take a long time using Markov Chian Monte Carlo (MCMC) optimisation techniques. Using \pkg{TMB}, the models are much faster and produce similar results.

>>>>

%% - \title{} in title case
%% - \Plaintitle{} without LaTeX markup (if any)
%% - \Shorttitle{} with LaTeX markup (if any), used as running title
\title{disaggregation: An \proglang{R} Package for Bayesian Spatial Disaggregation Modelling}
\Plaintitle{disaggregation: An R Package for Bayesian Spatial Disaggregation Modelling}
\Shorttitle{Disaggregation Modelling in \proglang{R}}

@@

Disaggregation modelling is unorthodox as the predictions are at a different scale to the response data, i.e. the number of rows in the covariate data is different to that of the response data. The spatial modelling software package, \pkg{INLA}~\cite{rue2009approximate} (integrated nested Laplace approximation), has been shown to very useful in a wide variety of circumstances, however it is not flexible enough for the unorthodox nature of the disaggregation problem except in the special case of the linear link function \cite{wilson2017pointless}. Disaggregation models can be implemented in \pkg{TMB}~\cite{tmb2016} (template model builder) with a lot of flexibility, however the data manipulation required to format the model objects and construct the model definition in C++ is non-trivial. The \pkg{disaggregation} package allows this process to be streamlined, to make it easy for the user to run disaggregation models at the expense of some flexibility.

@@

The \pkg{disaggregation} package is built on template model builder (\pkg{TMB})~\cite{tmb2016}, which is a tool for flexibly building complex models based on C++. \pkg{TMB} combines the packages CppAD~\cite{bell2012cppad}, for automatic differentiation in C++, Eigen~\cite{guennebaud2010eigen}, a C++ library for linear algebra, and CHOLMOD~\cite{chen2008algorithm}, a package for efficient computation of sparse matrices. The use of these packages allows an efficient implementation of the automatic Laplace approximation~\cite{skaug2006automatic} with exact derivatives which gives an approximation to the Bayesian posterior. \pkg{TMB} calculates first and second order derivatives of the objective function using automatic differentiation~\cite{griewank2008evaluating}.

@@

%
<<setuppaths, results='hide', show='hide'>>=
library(raster)

@@

%
<<prepare_data, warning = FALSE>>=
dis_data <- prepare_data(polygon_shapefile = shapes, 
                         covariate_rasters = covariate_stack, 
                         aggregation_raster = population_raster, 
                         mesh.args = list(max.edge = c(0.7, 8), 
                                          cut = 0.05, 
                                          offset = c(1, 2)),
                         id_var = 'ID_2',
                         response_var = 'inc',
                         na.action = TRUE,
                         ncores = 8)
@
%

@@

\section{Comparison with Markov chain Monte Carlo (MCMC)} \label{sec:comparison}

In this section we show performance comparisons between modelling using the Laplace approximation provided by the \pkg{disaggregation} package, based on \pkg{TMB}, and using MCMC. Given a function to evaluate the probability density of a distribution at any given point in parameter space, MCMC algorithms construct markov chains to generate samples from this distribution. These algorithms are often slow, particularly in high-dimensional settings, as it can take a long time to converge to and effectively sample from the stationary distribution. The density is evaluated at each step, so this problem is compounded when evaluating this density is computationally expensive. In contrast, the \pkg{disaggregation} package approximates uses a Laplace approximation to the posterior to generate posterior samples. This only requires the posterior to be maximised to find the posterior mode and therefore involves relatively few evaluations of the posterior density compared to MCMC techniques, although potentially at the expense of less accurate posterior samples. Here we compare the time and performance of the two techniques.

The model described in Section~\ref{sec:code} for malaria in Madagascar has been optimised using the \pkg{disaggregation} package. Here we fit the same model by running MCMC using the \pkg{tmbstan} package, with the NUTS (no-u-turn sampler) algorithm \cite{hoffman2014no} using four chains. It is important to note that this is a useful feature of the \pkg{disaggregation} package, to be able to create the \pkg{TMB} model object (using the \code{fit_model} function with one iteration) and pass it directly to \pkg{tmbstan}. The model is fitted by running the MCMC algorithm for 8000 iterations with 2000 of those as warmup, which took 94 hours. This number of iterations was chosen by running the MCMC algorithm repeatedly, starting at 1000 iterations, doubling the number of iterations each time until the value of the MCMC convergence statistic, $\hat{R}$, dropped below 1.05 for all model parameters. In contrast, fitting the model using the Laplace approximation via \pkg{TMB} within the \pkg{disaggregation} package took 56 seconds. Fitted parameter values for both of these methods are given in Table~\ref{tab:mcmcresults}.

@@

The modelling framework is implemented using the Laplace approximation and automatic differentiation within the \pkg{TMB} package. This allows fast, optimised calculations in C++. These disaggregation models are computationally intensive and take a long time using MCMC optimisation techniques. Using \pkg{TMB}, the models are much faster and produce similar results.
